---
title: "Download and preprocess the tonsil atlas data (Visium)"
author: "Ramon Massoni-Badosa, Marc Elosua-Bayes & Helena Lucia Crowell"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 100)
```

# Introduction

Here we will download and process the Seurat objects associated with the tonsil cell atlas to prepare them for submission to [`ExperimentHub`](https://bioconductor.org/packages/release/bioc/html/ExperimentHub.html). We will follow a similar style and structure as the one used for the [`TabulaMurisSenisData`](https://github.com/fmicompbio/TabulaMurisSenisData) pacakge.

In particular, here we will focus on the [Visium](https://www.10xgenomics.com/products/spatial-gene-expression) data from our tonsil atlas.

# Loading required packages

```{r dependencies}
library(here)
library(glue)
library(Seurat)
library(HDF5Array)
library(stringr)
library(SpatialExperiment)
library(tools)
library(dplyr)
library(readr)
```


# Downloading the Seurat objects from Zenodo

<!--
The Seurat objects derived from our tonsil atlas have been archived as .rds files in Zenodo under the record [8373756](https://zenodo.org/record/8373756). To generate the URL's for each dataset we used the [zenodo_get](https://github.com/dvolgyes/zenodo_get) package as follows:

```{bash}
zenodo_get 10.5281/zenodo.8373756 -w tmp.txt
paste tmp.txt md5sums.txt | sed 's/\t/,/' | sed 's/ \{1,\}/,/g' > tonsil-atlas-zenodo-get-output_v2.csv
```
-->

The chunk above was run already to download the scRNA-seq data in another notebook, so we can read the "tonsil-atlas-zenodo-get-output_v2.csv" file and get the URL of the Visium data:

```{r}
# Read data
path_to_csv1 <- here("inst/scripts/tonsil-atlas-zenodo-get-output_v2.csv")
files_df <- read.csv(file = path_to_csv1, header = FALSE)
colnames(files_df) <- c("url", "file_id", "file_name")
```

We proceed to download the Visium dataset:

```{r}
# Create data directories
data_dir <- here("inst/scripts/spatial_raw_data")
out_dir <- here("inst/scripts/HCATonsilData/2.0/Spatial")
dir.create(data_dir, recursive = TRUE, showWarnings = TRUE)
dir.create(out_dir, recursive = TRUE, showWarnings = TRUE)


# Get URL and download
options(timeout = 1000000)
filt <- str_detect(files_df$file_name, "Spatial")
spatial_url <- files_df[filt, "url"]
out_file <- file.path(
  data_dir,
  files_df[filt, "file_name"]
)
download.file(url = spatial_url, destfile = out_file)


# Check MD5
md5_original <- files_df[filt, "file_id"]
md5_download <- md5sum(out_file)
if (!md5_original == md5_download) {
  stop("The file is corrupted!")
}


# Uncompress
untar(out_file, exdir = data_dir)
```


# Process Seurat objects to extract relevant slots

```{r}
# Read data
file_path <- here("inst/scripts/spatial_raw_data/spatial_transcriptomics/20220527_tonsil_atlas_spatial_seurat_obj.rds")
so <- readRDS(file_path)


# 'assays'
for (i in c("counts", "data")) {
    j <- switch(i, counts="counts", data="logcounts")
    fnm <- file.path(out_dir, glue("Spatial_assay_{j}.h5"))
    mtx <- GetAssayData(so, i)
    h5 <- writeHDF5Array(mtx, 
        filepath=fnm, name=j, 
        chunkdim=getHDF5DumpChunkDim(dim(mtx)))
}

# 'reducedDims'
for (dr in Reductions(so)) {
    obj <- Embeddings(so, dr)
    rownames(obj) <- NULL
    fnm <- file.path(out_dir, glue("Spatial_dimred_{dr}.rds"))
    saveRDS(obj, fnm)
}


# 'colData'
cd <- DataFrame(so@meta.data)
cd$sample_id <- cd$gem_id # needed for SPE
saveRDS(cd, file.path(out_dir, "Spatial_coldata.rds"))


# 'rowData'
genes_df <- read_tsv(
  here("inst/extdata/features.tsv.gz"),
  col_names = c("ensembl_id", "symbol", "type")
) %>%
  filter(type == "Gene Expression") %>%
  as.data.frame()
gene_ids <- genes_df$ensembl_id
names(gene_ids) <- genes_df$symbol
gs <- rownames(so)
rd <- so[["Spatial"]]@meta.features
rd <- DataFrame(row.names=gs, gene_name=gs, rd)
rd$gene_id <- gene_ids[rd$gene_name]
names(rd$gene_id) <- NULL
rd$highly_variable <- gs %in% VariableFeatures(so)
saveRDS(rd, file.path(out_dir, "Spatial_rowdata.rds"))

# 'imgData'
id <- lapply(Images(so), \(id) {
    sfs <- so@images[[id]]@scale.factors
    img <- GetImage(so, image=id, mode="raster")
    saveRDS(img, file.path(out_dir, glue("Spatial_image_{id}.rds")))
    saveRDS(sfs, file.path(out_dir, glue("Spatial_scale_{id}.rds")))
})
```


# Sanity checks

As a sanity check, let us ensure that we can create successfully each SingleCellExperiment object from its individual slots:

```{r load}
# metadata
rd <- readRDS(file.path(out_dir, "Spatial_rowdata.rds"))
cd <- readRDS(file.path(out_dir, "Spatial_coldata.rds"))

# assays
as <- list.files(out_dir, "^Spatial_assay_", full.names=TRUE)
names(as) <- gsub("^Spatial_assay_(.*)\\.h5$", "\\1", basename(as))
as <- mapply(
    \(h5, nm) {
        mtx <- h5read(h5, name=nm)
        rownames(mtx) <- rd$gene_name
        colnames(mtx) <- cd$barcode
        as(mtx, "dgCMatrix")
    },
    h5=as, nm=names(as), SIMPLIFY=FALSE)

# reductions
dr <- list.files(out_dir, "^Spatial_dimred_", full.names=TRUE)
names(dr) <- toupper(gsub("^Spatial_dimred_(.*)\\.rds", "\\1", basename(dr)))
dr <- lapply(dr, readRDS)

# images
id <- lapply(unique(cd$sample_id), \(id) {
    img <- readRDS(file.path(out_dir, glue("Spatial_image_{id}.rds")))
    sfs <- readRDS(file.path(out_dir, glue("Spatial_scale_{id}.rds")))
    DataFrame(
        sample_id=id, image_id="lowres",
        data=I(list(SpatialImage(img))), 
        scaleFactor=sfs$lowres)
}) |> do.call(what="rbind")
```

Construct SpatialExperiment object:

```{r spe}
# construct 'SpatialExperiment'
(spe <- SpatialExperiment(assays=as, 
    mainExpName="Spatial",
    reducedDims=dr, imgData=id,
    rowData=rd, colData=cd)) #metadata=md
```

```{r check, warning=FALSE}
sce_old <- as.SingleCellExperiment(so)
sce_new <- as(spe, "SingleCellExperiment")
# drop class-specific identifiers
sce_old$ident <- sce_new$sample_id <- NULL
# drop 'spatialCoords'
int_colData(sce_new)$spatialCoords <- NULL
# match ordering of reduced dimensions
drs <- reducedDimNames(sce_old)
reducedDims(sce_new) <- reducedDims(sce_new)[drs]
# drop irrelevant internals & dimension names
imd <- c("version", "mainExpName")
int_metadata(sce_old) <- int_metadata(sce_old)[imd]
int_metadata(sce_new) <- int_metadata(sce_new)[imd]
rownames(colData(sce_old)) <- rownames(colData(sce_new)) <- NULL
# this is necessary, too...
elementMetadata(rowRanges(sce_old)) <- elementMetadata(rowRanges(sce_new))
# check saved & re-loaded objects match
identical(sce_old, sce_new)
```


# Session Information

```{r session-info}
sessionInfo()
```
